1-intro

### 2022-03-01 1-intro.Rmd
Update 1-intro.Rmd
minor polishing

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time intervals. This type of data can naturally be considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layout: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center'}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

In the \proglang{R} community, many data objects have been proposed to work with specific spatio-temporal tasks, while few can freely explore the data. \pkg{spacetime} [@spacetime] is an attempt for such a data object, but the spatial and temporal class it built on have now been gradually replaced by more recent implementations. The \pkg{stars} [@stars] package uses an array to wrangle spatio-temporal data, but the dense array it uses can quickly exhaust the memory for data in the irregular spatial grid. While there has not yet been a data object that makes it easy to wrangle spatio-temporal data, there is a strong need for such tools given the ubiquitous spatial and temporal data collected in the society, i.e., climate observations from weather stations, air quality variables from air monitoring stations. The demand for such tools can also come from spatial data analysts wishing to incorporate temporal data in their analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to overcome the problem with organising variables in different levels when working with spatio-temporal data. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With the new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section 2 presents the main design and functionality of \pkg{cubble}. Section 3 explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section 4 uses Australia weather station data and river level data as examples to demonstrate the use with \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section 5 concludes the paper.

>>>>

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time intervals. This type of data can naturally be considered as a data cube with three axes: spatial identifier, temporal identifier, and variables. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center'}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

In the \proglang{R} community, many data objects have been proposed to work with specific spatio-temporal tasks, while few can freely explore the data. \pkg{spacetime} [@spacetime] is an attempt for such a data object, but the spatial and temporal classes it is built on have now been gradually replaced by more recent implementations. The \pkg{stars} [@stars] package uses an array to wrangle spatio-temporal data, but the dense array it uses can quickly exhaust the memory for data in the case of irregular spatial grid. While there has not yet been a data object that makes it easy to wrangle spatio-temporal data, there is a strong need for such tools given the ubiquitous spatial and temporal data collected in the society, i.e., climate observations from weather stations, air quality variables from air monitoring stations. The demand for such tools can also come from spatial data analysts wishing to incorporate temporal data in their analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to overcome the problem with organising variables in different levels when working with spatio-temporal data. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section 2 presents the main design and functionality of \pkg{cubble}. Section 3 explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section 4 uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section 5 concludes the paper.
